### 2023-02-07 paper-thesis/chapters/02-background-spoofax.tex
Rewrite spoofax background

The meta-language SDF3 is used to define the syntax of the language~\cite{sdf3}, which is then parsed using Spoofax's SGLR parser~\cite{sdf3_parser}. The parser produces a abstract syntax tree in the ATerm format~\cite{aterm}. A pretty printer is also generated based on this syntax definition.

\paragraph{Syntax definitions}
The syntax of

\paragraph{ATerm format}

\section[Statix and Scope graphs]{\label{sec:statix}Statix and Scope graphs\protect\footnote{Note: This section is adapted from "Composable Type System Specification using Heterogeneous Scope Graphs".~\cite[sect. 4.1.2]{Zwaan21}}}

In Statix, scope graphs are used to represent a programs name and type information~\cite{scopes_as_types}~\cite{nameres}. A scope graph is a graph that consists of the following components:

>>>>

The meta-language SDF3 is used to define the syntax of the language~\cite{sdf3}, which is then parsed using Spoofax's SGLR parser~\cite{sdf3_parser}.

\paragraph{Syntax definitions}
The syntax of a language is defined through rules of the form \verb|A.C = a| with \verb|A| being a \emph{Non-Terminal}, \verb|C| being the name of a constructor and \verb|a| a list of symbols. To remove ambiguity, one can specify a relative priority for each constructor. For example, below is the syntax of a simple language:
\begin{lstlisting}
context-free sorts
	Expr
context-free syntax
	Expr.Add = Expr "+" Expr
	Expr.Mul = Expr "*" Expr
	Expr.X = "x"
context-free priorities
	Expr.X > Expr.Mul > Expr.Add
\end{lstlisting}

\paragraph{ATerm format}
Using the Syntax, Spoofax will generate a parser. This parser takes in a program in textual format, and outputs ATerms. ATerms are a way of encoding an abstract syntax tree~\cite{aterm}. For example, for the program \verb|x + x * x| the following ATerm is produced:
\begin{lstlisting}
Add(X(), Mul(X(), X()))
\end{lstlisting}

\section{Statix}
Now that the syntax has been determined, we move on to the static semantics. Statix is a declarative language for describing the static semantics of a language~\cite{scopes_as_types}. Using these rules, Statix automatically derives a type checker. The rules are based on \emph{equality constraints}. The basic constraints
are equality constraints \verb|t1 == t2| and inequality constraints \verb|t1 != t2| on terms. An equality constraint is states that two terms are equal or can be unified. An inequality constraint states that two terms are not equal.

The constraints are generated by predicates. A predicate defines a relation on terms and scopes, it is of the form $\mathtt{p(t_1, ..., t_n) :- c_1, ..., c_n}$. Each constraint $c_1...c_n$ is another predicate or an equality constraint. Below is an example that uses these mechanics to check that an expression does not contain any \verb|Muls|. Note that \verb|noMuls| is not defined for the \verb|Mul| constructor, so it will fail if it encounters one.
\begin{lstlisting}
noMuls(Add(e1, e2)) :- noMuls(e1), noMuls(e2).
noMuls(X()).
\end{lstlisting}

Statix also has functional predicates, these define a one-to-one relationship between input terms and the output term. For example below we use this to associate a type with each expression.
\begin{lstlisting}
typeOfExpr(X()) = INT().
typeOfExpr(Add(e1, e2)) = INT() :-
	typeOfExpr(e1) == INT(),
	typeOfExpr(e2) == INT().
\end{lstlisting}

In Statix, the language designer writes down the declarative semantics. The Statix solver decides the order in which constraints are considered, the language designer has no control over that.

\section[Scope graphs]{\label{sec:statix}Scope graphs\protect\footnote{Note: This section is adapted from "Composable Type System Specification using Heterogeneous Scope Graphs".~\cite[sect. 4.1.2]{Zwaan21}}}

Statix also has scope graphs. Scope graphs are used to represent a programs' name and type information~\cite{scopes_as_types}~\cite{nameres}. A scope graph is a graph that consists of the following components:
